\chapter{Compiling \Stan Programs}

\noindent
Preparing a \Stan program to be run involves two steps,
%
\begin{enumerate}
\item translating the \Stan program to \Cpp, and
\item compiling the resulting \Cpp to an executable.
\end{enumerate}
%
This chapter discusses both steps, as well as their encapsulation into
a single make target.  

\section{Installing Stan}

Before Stan can be run, it must be installed; see
\refappendix{install} for complete platform-specific installation
details.

\section{Translating and Compiling through {\tt\bfseries make}}\label{make-models.section}

The simplest way to compile a Stan program is through the \code{make}
build tool, which encapsulates the translation and compilation step
into a single command.  The commands making up the \code{make} target
for compiling a model are described in the following sections, and the
following chapter describes how to run a compiled model.

\subsection{Translating and Compiling Test Models}

There are a number of test models distributed with Stan which unpack
into the path \code{src/models}.  To build the simple example
\nolinkurl{src/models/basic\_estimators/bernoulli.stan}, the following call
to \code{make} suffices.  First the directory is changed to Stan's home
directory by replacing \code{<stan-home>} with the appropriate path.
%
\begin{quote}
\begin{Verbatim}[fontshape=sl]
> cd <stan-home>
\end{Verbatim}
\end{quote}
%
The current directory should now contain the file named
\code{makefile}, which is the default instructions used by
\code{make}.  From within the top-level Stan directory, the following call
will build an executable form of the Bernoulli estimator.
%
\begin{quote}
\begin{Verbatim}[fontshape=sl]
> make src/models/basic_estimators/bernoulli
\end{Verbatim}
\end{quote}
%
This will translate the model \code{bernoulli.stan} to a \Cpp file and
compile that \Cpp file, putting the executable in
\nolinkurl{src/models/basic_distributions/bernoulli(.exe)}.  Although
the make command including arguments is itself portable, the target it
creates is different under Windows than in Unix-like platforms.  Under
Linux and the Mac, the executable will be called \code{bernoulli},
whereas under Windows it will be called \code{bernoulli.exe}.

\subsection{Dependencies in {\tt\bfseries make}}

A \code{make} target can depend on other \code{make} targets.  When
executing a \code{make} target, first all of the targets on which it
depends are checked to see if they are up to date, and if they are
not, they are rebuilt.  This includes the top-level target itself.  If
the \code{make} target to build the Bernoulli estimator is invoked a
second time, it will see that it is up to date, and not compile
anything.  But if one of the underlying files has changes since the
last invocation \code{make}, such as the model specification file, it
will be retranslated to \Cpp and recompiled to an executable.

There is a dependency included in the make target that will
automatically build the \code{bin/stanc} compiler and the
\code{bin/libstan.a} library whenever building a model.


%\subsection{Making Other Models}
%
%A \code{make} target can be invoked from a directory other than the
%one that contains the \code{makefile} with the \code{-f} option.  Thus
%an alternative way to compile this model from its own source directory
%would be as follows (again replacing \code{<stan-home>} with the
%appropriate path to the top-level directory in which Stan was
%unpacked).
%\begin{quote}
%\begin{Verbatim}[fontshape=sl]
%> cd <stan-home>src/models/basic_estimators
%> make -f ../../../makefile bernoulli
%\end{Verbatim}
%\end{quote}
%%
%The expression \code{../../../makefile} is a relative path to the
%file named \code{makefile}; this says to go up three directories
%(i.e., back to \code{<stan-home>}) and then look for a file named
%\code{makefile}.  This can be an absolute path, too.  The Unix-like
%forward slashes can be used under Windows as well as under Linux and
%Mac.  


\subsection{Getting Help from the {\tt makefile}}

Stan's \code{makefile}, which contains the top-level instructions to
\code{make}, provides extensive help in terms of targets and options.
It is located at the top-level of the distribution, so first change
directories to that location.
%
\begin{quote}
\begin{Verbatim}[fontshape=sl]
> cd <stan-home>
\end{Verbatim}
\end{quote} 
and then invoke make with the target \code{help},

\begin{quote}
\begin{Verbatim}[fontshape=sl]
> make help
\end{Verbatim}
\end{quote}


\subsection{Options to \code{make}}

Stan's \code{make} targets allow the user to change compilers,
library versions for Eigen and Boost, as well as compilation options
such as optimization.

These options should be placed right after the call to \code{make}
itself.  For instance, to specify the \code{clang++} compiler at
optimization level 0, use
%
\begin{quote}
\begin{Verbatim}[fontshape=sl]
> make CC=clang++ O=0 ...
\end{Verbatim}
\end{quote}


\subsubsection{Compiler Option}

The option \code{CC=g++} specifies the \code{g++} compiler and
\code{CC=clang++} specifies the \code{clang++} compiler.  Other
compilers with other names may be specified the same way.  A full path
may be used, or just the name of the program if it can be found on the
system execution path.  

\subsubsection{Optimization Option}

The option \code{O=0} (that's letter `O', equal sign, digit `0'),
specifies optimization level 0 (no optimization), whereas \code{O=3}
specifies optimization level 3 (effectively full optimization), with
levels 1 and 2 in between.

With higher optimization levels, generated executable tends to be bigger
(in terms of bytes in memory) and faster.  For best results on
computationally-intensive models, use optimization level 3 for the
Stan library and for compiling models.

\subsubsection{Library Options}

Alternative versions of Eigen, Boost, and Google Test may be specified
using the properties \code{EIGEN}, \code{BOOST}, and \code{GTEST}.
Just set them equal to a path that resolves to an appropriate library.
See the libraries distributed under \code{lib} to see which
subdirectory of the library distribution should be specified in order
for the include paths in the \Cpp code to resolve properly.


\subsection{Additional  \code{make} Targets}

All of these targets are intended to be invoked from the top-level
directory in which Stan was unpacked (i.e., the directory that
contains the file named \code{makefile}).

\subsubsection{Clean Targets}

A very useful target is \code{clean-all}, invoked as
%
\begin{quote}
\begin{Verbatim}[fontshape=sl]
> make clean-all
\end{Verbatim}
\end{quote}
%
This removes everything that's created automatically by \code{make},
including the \code{stanc} translator, the Stan libraries, and all the
automatically generated documentation.  

\subsubsection{Make Target for \code{stanc}}

To make the \code{stanc} compiler, use
%
\begin{quote}
\begin{Verbatim}[fontshape=sl]
> make bin/stanc
\end{Verbatim}
\end{quote}
%
As with other executables, the executable \code{bin/stanc} will be
created under Linux and Mac, whereas \code{bin/stanc.exe} will be
created under Windows.

\subsubsection{Make Target for Stan Library}

To build the Stan library, use the following target,
%
\begin{quote}
\begin{Verbatim}[fontshape=sl]
> make bin/libstan.a
\end{Verbatim}
\end{quote}



\section{Translating \Stan to C++ with {\tt\bfseries stanc}}\label{stanc.section}


\subsection{Building the \stanc Compiler and the Stan Library}

Before the \stanc compiler can be used, it must be built.  Use the
following command from the top-level distribution directory containing
the file named \code{makefile}.
%
\begin{quote}
\begin{Verbatim}[fontshape=sl]
> make bin/stanc
\end{Verbatim}
\end{quote}
%
This invocation produces the executable \code{bin/stanc} under Linux
and Mac, and \code{bin/stanc.exe} under Windows.  The invocation of
\code{make}, including the forward slash, is the same on both platforms.

The default compiler option is \code{CC=g++} and the default
optimization level is \code{O=3} (the letter `O');  to see how to
change these, see the previous section in this chapter on \code{make}.


\subsection{The \stanc Compiler}

The \stanc compiler converts \Stan programs to \Cpp programs.  The
first stage of compilation involves parsing the text of the \Stan
program.  If the parser is successful, the second stage of compilation
generates \Cpp code.  If the parser fails, it will provide a
diagnostic error message indicating the location in the input where
the failure occurred and reason for the failure.

The following example illustrates a fully qualified call to \stanc
to build the simple Bernoulli model; just replace \code{<stan-home>}
with the top-level directory containing Stan (i.e., the directory
containing the file named \code{makefile}). 

For Linux and Mac:
%
\begin{quote}
\begin{Verbatim}[fontshape=sl]
> cd <stan-home>
> bin/stanc --name=bernoulli --o=bernoulli.cpp \
  src/models/basic_estimators/bernoulli.stan 
\end{Verbatim}
\end{quote}
%
The backslash (\Verb|\|) indicates a continuation of the same line.

For Windows:
%
\begin{quote}
\begin{Verbatim}[fontshape=sl]
> cd <stan-home>
> bin\stanc --name=bernoulli --o=bernoulli.cpp ^
    src/models/basic_estimators/bernoulli.stan 
\end{Verbatim}
\end{quote}
%
The caret (\Verb|^|) indicates continuation on Windows.

This call specifies the name of the model, here {\tt bernoulli}.
This will determine the name of the class implementing the model in
the \Cpp code.  Because this name is the name of a \Cpp class, it must
start with an alphabetic character (\code{a--z} or \code{A--Z}) and
contain only alphanumeric characters (\code{a--z}, \code{A--Z}, and
\code{0--9}) and underscores (\code{\_}) and should not conflict with
any \Cpp reserved keyword.  

The \Cpp code implementing the class is written to the file
\code{bernoulli.cpp} in the current directory.  The final argument,
\code{bernoulli.stan}, is the file from which to read the \Stan
program.

\subsection{Command-Line Options for {\tt\bfseries stanc}}

The model translation program \code{stanc} is called as follows.
%
\begin{quote}
\code{> stanc [options] {\slshape model\_file}}
\end{quote}
%
The argument \code{\slshape model\_file} is a path to a Stan model file ending
in suffix \code{.stan}.  The options are as follows.
%
\begin{description}
%
\item[\tt {-}-help] 
\mbox{ } \\ 
Displays the manual page for \stanc.  If this option is selected,
nothing else is done.
%
\item[\tt {-}-version]
\mbox{ } \\ 
Prints the version of \stanc.  This is useful for bug reporting
and asking for help on the mailing lists.
%
\item[\tt {-}-name={\slshape class\_name}]
\mbox{ } \\ 
Specify the name of the class used for the implementation of the
\Stan model in the generated \Cpp code.  
\\[2pt]
Default: {\tt {\slshape class\_name = model\_file}\_model}
%
\item[\tt {-}-o={\slshape cpp\_file\_name}]
\mbox{ } \\ 
Specify the name of the file into which the generated \Cpp is written.
\\[2pt]
Default: {\tt {\slshape cpp\_file\_name} = {\slshape class\_name}.cpp}
%
\item[\tt {-}-no\_main]
\mbox{ } \\
Include this flag to prevent the generation of a main function in the
output.
\\[2pt]
Default: generate a main function
\end{description}



\section{Compiling C++ Programs}\label{compiling-cpp.section}

\noindent
As shown in the previous section (\refsection{stanc}), \Stan converts
a program in the \Stan modeling language to a \Cpp program.  This \Cpp
program must then be compiled using a \Cpp compiler.  

The \Cpp compilation step described in this chapter, the model
translation step described in the last chapter, and the compilation of
the dependent binaries \code{bin/stanc} and \code{bin/libstan.a} may
be automated through make; see \refsection{make-models} for details.

\subsection{Which Compiler?}

\Stan has been developed using two portable, open-source \Cpp
compilers, \gpp and \clang, both of which run under and generate code
for Windows, Macintosh, and Unix/Linux.%
%
\footnote{As of the current version, \Stan cannot be compiled using
  \MSVC, the Windows-specific compiler from Microsoft.  \MSVC is able
  to compile the \code{stanc} compiler, but not the templates required
  for algorithmic differentiation and the Eigen matrix library.}

The \clang compiler is almost twice as fast at low levels of
optimization, but the machine code generated by \gpp at high
optimization levels is faster.


\subsection{What the Compiler Does}

A \Cpp compiler like \gpp or \clang performs several lower-level
operations in sequence,
% 
\begin{enumerate}
\item
parsing the input \Cpp source file(s), 
\item 
generating (static or dynamically) relocatable object code, and
\item 
linking the relocatable object code into executable code.
\end{enumerate}
%
These stages may be called separately, though the examples in this
manual perform them in a single call.  The compiler invokes the
assembler to convert assembly language code to machine code, and the
linker to resolve the location of references in the relocatable object
files.

\subsection{Compiler Optimization}

\Stan was written with an optimizing compiler in mind, which allows
the code to be kept relatively clean and modular.  As a result, \Stan
code runs as much as an order of magnitude or more faster with
optimization turned on.

For development of \Cpp code for \Stan, use optimization level 0; for
sampling, use optimization level 3.  These are controlled through
\Stan's makefile using \code{O=0} and directly through \clang or \gpp
with \code{-O0}; in both cases, the first character is the letter `O'
and the second the digit `0'.


\subsection{Building the \Stan Library}

Before compiling a \Stan-generated \Cpp program, the \Stan object
library archive must be built using the makefile.  This only needs to
be done once and then the archive may be reused.  The recommended
build command for the \Stan archive is as follows (replacing
\code{<stan-home>} with the directory into which Stan was unpacked and
which contains the file named \code{makefile}).
%
\begin{quote}
\begin{Verbatim}[fontshape=sl]
> cd <stan-home>
> make CC=g++ O=3 bin/libstan.a 
\end{Verbatim}
\end{quote}
%
Please be patient and ignore the (unused function) warning messages.
Compilation with high optimization on \code{g++} takes time (as much
as 10 minutes or more) and memory (as much as 3GB). 

This example uses the \code{g++} compiler for \Cpp (makefile option
\code{CC=g++}).  The \clang compiler may be used by specifying
\code{CC=clang++}.

This example uses compiler optimization level 3 (makefile option
\code{O=3}).  Turning the optimization level down to 0 allows the code
to built in under a minute in less than 1GB of memory.  This will slow
down sampling as much as an order of magnitude or more, so it is not
recommended for running models.  It can be useful for working on
\Stan's \Cpp code.


\subsection{Compiling a \Stan Model}

Suppose following the instructions in the last chapter
(\refsection{stanc}) that a \Stan program has been converted to a \Cpp
program that resides in the source file \code{<stan-home>/my\_model.cpp}.

The following commands will produce an executable in the file
\code{my\_model} in the current working directory (\code{<stan-home>}).
%
\begin{quote}
\begin{Verbatim}[fontshape=sl]
> cd <stan-home>
> g++ -O3 -Lbin -Isrc -isystem lib/boost_1.54.0              \
    -isystem lib/eigen_3.2.0 my_model.cpp -o my_model -lstan
\end{Verbatim}
\end{quote} %
The backslash (\Verb|\|) is used to indicate that the command is
continued; it should be entered all one one line.
%
The options used here are as follows.
\begin{quote}
\begin{description}
\item[\code{-O3}] sets optimization level 3,
\item[\code{-Lbin}] specifies that the archive is in the \code{bin}
  directory,
\item[\code{-Isrc}] specifies that the directory \code{src} should be
  searched for code (it contains the top-level \Stan headers),
\item[\code{-isystem lib/boost\_1.54.0}] specifies the include directory for the 
  Boost library,
\item[\code{-isystem lib/eigen\_3.2.0}] specifies the include directory for
  the Eigen library,
\item[\code{my\_model.cpp}] specifies the name of the source file to
  compile, and 
\item[\code{-o my\_model}] is the name of the resulting executable
  produced by the command (suffixed by \code{.exe} in Windows).
\item[\code{-lstan}] specifies the name of the archived library (not
  the name of the file in which it resides),
\end{description}
\end{quote}
%
The library binary and source specifications are required, as is the
name of the \Cpp file to compile.  User-supplied directories may be
included in header or archive form by specifying additional \code{-L},
\code{-l}, and \code{-I} options.
 
A lower optimization level may be specified.  If there is no
executable name specified using the \code{-o} option, then the model
is written into a file named \code{a.out}.
 

\subsection{Library Dependencies}

\Stan depends on two open-source libraries,
%
\begin{enumerate}
\item the Boost general purpose \Cpp libraries, and 
\item the Eigen matrix and linear algebra \Cpp libraries.
\end{enumerate}
%
These are both distributed along with \Stan in the directory
\code{<stan-home>/lib}.  

The code for \Stan itself is located in the directory
\code{<stan-home>/src}.  Because not all of \Stan is included in the
archive \code{bin/libstan.a}, the \code{src} directory must also be
included for compilation.




