\chapter{Licensing}\label{licensing.appendix}

\noindent
\CmdStan, Stan, and Stan's three dependent libraries, Stan Math Library, Boost, and Eigen, are
distributed under liberal freedom-respecting%
%
\footnote{The link
  \url{http://www.gnu.org/philosophy/open-source-misses-the-point.html}
  leads to a discussion about terms ``open
  source'' and ``freedom respecting.''}
%
licenses approved by the Open Source Initiative.%
\footnote{See \url{http://opensource.org}.}  

In particular, the licenses for \CmdStan and its dependent libraries
have no ``copyleft'' provisions requiring applications of \CmdStan to
be open source if they are redistributed.

This chapter describes the licenses for the tools that are distributed
with \CmdStan.  The next chapter explains some of the build tools that
are not distributed with \CmdStan, but are required to build and run
Stan models.

\section{\CmdStan's License}

\CmdStan is distributed under the BSD 3-clause license (BSD New).
%
\begin{quote}
\url{http://www.opensource.org/licenses/BSD-3-Clause}
\end{quote}

\section{Stan's License}

Stan is distributed under the BSD 3-clause license (BSD New).
%
\begin{quote}
\url{http://www.opensource.org/licenses/BSD-3-Clause}
\end{quote}

\section{Stan Math Library's License}

The Stan Math Library is distributed under the BSD 3-clause license (BSD New).
%
\begin{quote}
\url{http://www.opensource.org/licenses/BSD-3-Clause}
\end{quote}

\section{Boost License}

Boost is distributed under the Boost Software
License version 1.0.
%
\begin{quote}
\url{http://www.opensource.org/licenses/BSL-1.0}
\end{quote}

\section{Eigen License} 
%
Eigen is distributed under the Mozilla Public License,
version 2.
%
\begin{quote}
\url{http:/http://opensource.org/licenses/mpl-2.0}
\end{quote}

\section{CVODES License} 
%
CVODES is distributed under the BSD 3-clause license (BSD New).
%
\begin{quote}
\url{http://computation.llnl.gov/projects/sundials-suite-nonlinear-differential-algebraic-equation-solvers/license}
\end{quote}

\section{Google Test License}

\CmdStan uses Google Test for unit testing; it is not required to
compile or execute models.  Google Test is distributed under the BSD
2-clause license.
%
\begin{quote}
\url{http://www.opensource.org/licenses/BSD-License}
\end{quote}

